---
title: "Class 11: Structural Bioinformatics"
date: 11/6/2018
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#1.1 PDB Statistics

Download PDB statictics as CSV file from  http://www.rcsb.org/stats/summary
```{r}
pdbstats <- read.csv("Data Export Summary.csv", row.names = 1)
```

lets look at the table
```{r}
library(knitr)
kable(pdbstats)
```

Q1: Determine the percentage of structures solved by X-Ray and Electron Microscopy. What proportion of structures are protein?
```{r}
sumTotal <- sum(pdbstats$Total)
percentage <- round(pdbstats$Total / sumTotal*100, 2)
```

A1: The percentage of structures solved by X-ray is `r percentage[1]`% and the perentage of structures solved by electron microscopy is `r percentage[3]`% in the protein database as `r Sys.Date()`

adding the percentage of each method as a new column of the table 
```{r}
nstats <- pdbstats
nstats$Percent <- percentage
kable(nstats)
```

Calculate the proportion of protein entries
```{r}
protein <- sum(pdbstats$Proteins)
proportionP <- round(protein / sumTotal*100, 2)
```
A2: The proportion of protein entries from the summary table is `r proportionP`%

# Introduction to Bio3D

Load the bio3d library
```{r}
library(bio3d)
```

read the pdb file 
```{r}
pdb <- read.pdb(file ="1hsg.pdb")
attributes(pdb)
```
Q8. Use the Bio3D   function to write out a protein only PDB file for viewing in VMD. Also write out a second separate PDB file for the ligand with residue name MK1
```{r}
ligand.inds <- atom.select(pdb,"ligand", value = TRUE)
write.pdb(pdb, "ligand.pdb", type = pdb$ligand.inds)
```

#Align multiple structures 
```{r}
#Download some example pdb files
ids <- c("1TND_B","1AGR_A","1TAG_A","1GG2_A","1KJY_A","4G5Q_A")
files <- get.pdb(ids, split = TRUE)
```

align these strucutures
```{r}
# Extract and align the chains we are interested in
pdbs <- pdbaln(files, fit = TRUE)
# Print to screen a summary of the 'pdbs' object
pdbs
```

Do PCA analysis 
```{r}
# find invariant core
core <- core.find(pdbs)
# superimpose all structures to core
pdbs$xyz = pdbfit(pdbs, core)
# Perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)
```

Function rmsd() will calculate all pairwise RMSD values of the structural ensemble. 
This facilitates clustering analysis based on the pairwise structural deviation:
```{r}
# Calculate RMSD
rd <- rmsd(pdbs)
# Structure-based clustering
hc.rd <- hclust(dist(rd)) 
grps.rd <- cutree(hc.rd, k=3)
plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```

To visualize the major structural variations in the ensemble the function mktrj() can be used to generate a trajectory PDB file by interpolating along a give PC (eigenvector):
```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```

