---
#header of the file 
title: "Class 8"
date: 10/25/2018
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-means clustering

Our first example with **kmeans()** function.

```{r}
# Example plot to see how Rmarkdown works
# Shortcut for insert code chunks: option+command+i
plot(1:10, typ = "l")
```

Back to kmeans
```{r}
# Generate some example data for clustering
tmp <- c(rnorm(30,-3), rnorm(30,3))  # this is a vector, c is used to concatenate
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```

Use the kmeans() function setting k to 2 and nstart=20
```{r}
k <- kmeans(x, centers = 2, nstart = 20)
```

Inspect/print the results
```{r}
print(k)
```

Q. How many points are in each cluster?
```{r}
k$size
```
   - there are 30 points in each cluster

Q. What ‘component’ of your result object details
    - cluster size: `k$size`
    - cluster assignment/membership: clustering vector 
```{r}
table(k$cluster)
```
    - cluster center: values in cluster means
```{r}
k$centers
```
    - Within cluster sum of squares by cluster: square distance to the center
    
Plot x colored by the kmeans cluster assignment and add cluster centers as blue points
```{r}
palette(c("blue", "black"))
plot(x, col = k$cluster)
points(k$centers, col = "red", pch = 20, cex = 3)
```

Q. Repeat for k=3, which has the lower tot.withinss?
```{r}
k3 <- kmeans(x, centers = 3, nstart = 20)
k4 <- kmeans(x, centers = 4, nstart = 20)
k10 <- kmeans(x, centers = 10, nstart = 20)
k10$tot.withinss
k4$tot.withinss
k3$tot.withinss
k$tot.withinss
plot(c(k$tot.withinss, k3$tot.withinss, k4$tot.withinss, k10$tot.withinss))
```
 
  - k=3 has a lower tot.withinss (larger k will result lower tot.withinss)
Major disadvantage is that we have to give a k value each time.       



# Hierarchical clustering in R

Let's try out the **hclust()** function for Hierarchical clustering in R. This function need a **distance matrix** as input.

```{r}
d <- dist(x)
hc <- hclust(d)
#this plots a dendrogram 
plot(hc)
# Draw a line on the dendrogram
abline(h = 8, col = "red")
# Cut the tree to yeild cluster membership vector
cutree(hc, h = 8)  #cut by height
cutree(hc, k = 2)  #cut by k groups
```

Using different hierarchical clustering methods
```{r}
hc.complete <- hclust(d, method="complete")
hc.average <- hclust(d, method="average")
hc.single <- hclust(d, method="single")
```

### A more real example of hierarchical clustering
```{r}
### Step 1. Generate some example data for clustering
x <- rbind(
matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
# Step 2. Plot the data without clustering
plot(x)
# Step 3. Generate colors for known clusters
# (just so we can compare to hclust results)
palette(c("red","blue","black"))
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
plot(x, col=col, pch = 20)
```

Q. Use the dist(), hclust(), plot() and cutree() functions to return 2 and 3 clusters
```{r}
disX <- dist(x)
hcx <- hclust(disX)
plot(hcx)
cluster2 <- cutree(hcx, k =2)  #return 2 clusters
cluster3 <- cutree(hcx, k =3)  #return 3 clusters
palette(c("red","blue", "yellow"))
plot(x, col = cluster2, pch = 20)
plot(x, col = cluster3, pch = 20)
```

Q. How does this compare to your known 'col' groups?
```{r}
table(cluster3, col)
```
                   
                   

# Using principal component analysis (PCA) in R   

Generate a gene expression table 
```{r}
# Initialize a blank 100 row by 10 column matrix
mydata <- matrix(nrow=100, ncol=10)
# Lets label the rows gene1, gene2 etc. to gene100
rownames(mydata) <- paste("gene", 1:100, sep="")
# Lets label the first 5 columns from wt1 to wt5 and the last 5 fromko1 to ko5
colnames(mydata) <- c( paste("wt", 1:5, sep=""),
paste("ko", 1:5, sep="") )
# Fill in some fake read counts
for(i in 1:nrow(mydata)) {
wt.values <- rpois(5, lambda=sample(x=10:1000, size=1))
ko.values <- rpois(5, lambda=sample(x=10:1000, size=1))
mydata[i,] <- c(wt.values, ko.values)
}
head(mydata)
```

Note the prcomp() function wants us to take the transposed data of our matrix with column and rows flipped 
```{r}
head(t(mydata))
```

Now let's try find structure in this data with **prcomp()**
```{r}
pca <- prcomp(t(mydata), scale = TRUE)
#see what's returned by prcomp using attributes()
attributes(pca)
```

Make a PC plot of PC1 vs PC2, using `$x` component of output (pca$x)
```{r}
plot(pca$x[,1], pca$x[,2])
```



Looking the variance from the original data that we are capturing
```{r}
#variance: squared stdev
pca.var <- pca$sdev^2
pca.var
#getting the percentage variance 
pca.var.per <- round(pca.var/sum(pca.var)*100,1)
pca.var.per
```

Making the basic PCA plot more useful
```{r}
## A vector of colors for wt and ko samples
colvec <- colnames(mydata)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
xlab=paste0("PC1 (", pca.var.per[1], "%)"),
ylab=paste0("PC2 (", pca.var.per[2], "%)"))
```

Making a "Scree-plot" 
```{r}
barplot(pca.var.per, main="Scree Plot",
xlab="Principal Component", ylab="Percent Variation")
```

Looking at the loading scores
```{r}
# Lets focus on PC1 as it accounts for > 90% of variance
loading_scores <- pca$rotation[,1]
summary(loading_scores)
# We are interested in the magnitudes of both plus and minus contributing genes
gene_scores <- abs(loading_scores)
# Sort by magnitudes from high to low
gene_score_ranked <- sort(gene_scores, decreasing=TRUE)
# Find the names of the top 5 genes
top_5_genes <- names(gene_score_ranked[1:5])
# Show the scores (with +/- sign)
pca$rotation[top_5_genes,1]
```

# Try using what learned today to analyze UK food data
First read the food data and its dimension
```{r}
x <- read.csv("UK_foods.csv")
#reformat the row names
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
#check the dimension of x 
dim(x)
```

Read the whole data
```{r}
knitr::kable(x, caption="The full UK foods data table")
```

Plot a heatmap from this data set
```{r}
#par(mar=c(20, 4, 4, 2))
heatmap(as.matrix(x))
```

#Functions covered today
kmeans(x)
hclust(dist(x))
cutree(hc)
prcomp(x, scale = TRUE)
table()


