---
title: "Class13: Structural Bioinformatics II"
date: 11/13/2018
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1: Obtain and Inspect the HIV protein structure
Now load the Bio3D package and use the get.pdb() function to download the "1hsg" PDB entry into your RStudio Project directory:
```{r}
library(bio3d)
file.name <- get.pdb("1hsg")
```

Next use the read.pdb()function to read this PDB file into R so we can prepare it for further analysis.
```{r}
hiv <- read.pdb(file.name)
hiv
```

## Separate protein and ligand from the initial pdb file
We'll use `trim.pdb()`function to split our input structure into protein and ligand 
```{r}
prot <- trim.pdb(hiv, "protein")
lig  <- trim.pdb(hiv, "ligand")
write.pdb(prot, file="1hsg_protein.pdb")
write.pdb(lig, file="1hsg_ligand.pdb")
```

unix commands:
ls for list files in the current directory
ls ~ for list files in home directory
mv for move files 

# Section 2: Docking the ligand and inspect docking results
Inspect the docking result
```{r}
library(bio3d)
res <- read.pdb("all.pdbqt", multi=TRUE)
write.pdb(res, "results.pdb")
```

