---
title: "Class 16: Genome Analysis"
date: 11/27/2018
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read the input data

```{r}
metaFile <- "data/GSE37704_metadata.csv"
countFile <- "data/GSE37704_featurecounts.csv"
colData = read.csv(metaFile, row.names=1)
countData = read.csv(countFile, row.names=1)
head(colData)
head(countData)
```

remove the first column of countData

```{r}
countData <- as.matrix(countData[, -1]) #the minus sign means "everything but xxx"
head(countData)
```

Filter count data where you have 0 read count across all samples.
```{r}
countData <- countData[rowSums(countData)>0, ]
head(countData)
```

## Running DESeq2

Nice now lets setup the DESeqDataSet object required for the DESeq() function and then run the DESeq pipeline. This is again similar to our last days hands-on session.
```{r}
library(DESeq2)
dds = DESeqDataSetFromMatrix(countData=countData, colData=colData,design=~condition)
dds <- DESeq(dds)
```

See the results
```{r}
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))
summary(res)
```

## Plotting a volcano plot

Simple black plot
```{r}
plot( res$log2FoldChange, -log(res$padj))
```

Improve the quality of the plot
```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res))

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2)
mycols[inds] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col= mycols, 
      xlab="Log2(FoldChange)", ylab="-Log(P-value)")
```

## Add annotation to the gene data

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name = mapIds(org.Hs.eg.db,
                  keys=row.names(res),
                  keytype="ENSEMBL",
                  column="GENENAME",
                  multiVals="first")
head(res, 10)
```

# Section 2: Pathway Analysis

Load the packages:
```{r}
library(pathview)
library(gage)
library(gageData)
```

load data from kegg focusing on signalling and metabolism
```{r}
data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```

The main gage() function requires a named vector of fold changes, where the names of the values are the Entrez gene IDs.

Note that we used the mapIDs() function above to obtain Entrez gene IDs (stored in **res$entrez**) and we have the fold change results from DESeq2 analysis (stored in **res$log2FoldChange**).
```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

Now, let’s run the gage pathway analysis.

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
attributes(keggres)
head(keggres$less)
```

Use the pathview function

```{r}
# the pathway.id is from the results 
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

Now, let’s process our results a bit more to automagicaly pull out the top 5 upregulated pathways, then further process that just to get the pathway IDs needed by the pathview() function. We’ll use these KEGG pathway IDs for pathview plotting below.
```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

Finally, lets pass these IDs in keggresids to the pathview() function to draw plots for all the top 5 pathways.
```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

# Section 3: Gene Ontology (GO)

We can also do a similar procedure with gene ontology. Similar to above, go.sets.hs has all GO terms. go.subs.hs is a named list containing indexes for the BP, CC, and MF ontologies. Let’s focus on BP (a.k.a Biological Process) here.

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```

```{r}
head(gobpres$less)
```

#Section 4: Reactome Analysis
```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, 
            col.names=FALSE, quote=FALSE)
```

